---
title: "Exploratory Data Analysis Course Notes"
output:
  html_document:
    toc: yes
---  
## Principle of Analytic Graphics
* **Principle 1: Show Comparisons**
    * always comparative (compared to what)
	  * randomized trial - compare control group to test group
	  * evidence for a hypothesis is always relative to another competing hypothesis
    * ***example***: Two graphs showing the difference in Pokemon happiness after being taken on walks (graph 1: test) and not being taken on walks (graph 2: control)
  
* **Principle 2: Show explanation / causality**
    * what is the reason for the results shown by the test group?
    * form hypothesis to evidence showing a relationship (why something happened)
    * ***example: *** Another graph showing the level of Attention From Trainer for Pokemon who go on waks and those who do not. The difference in Pokemon Happiness within the Test Group could be explained by the increase in attention from the owner when Pokemon are taken on walks

* **Principle 3: Show multivariate data**
  * more than 2 variables because the real world is multivariate
	* show as much data on a plot as you can
	*   ***example***

```{r fig.height = 3, fig.width = 4, fig.align='center', echo=FALSE, warning=FALSE}
# install grid and png packages if not present
library(png)
library(grid)
grid.raster(readPNG("figures/1.jpg"))
```

* slightly negative relationship between pollution and mortality
	* when split up by season, the relationships are all positive $\rightarrow$ season = confounding variable

```{r fig.height = 4, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("figures/2.jpg"))
```


* **Principle 4: Integration of evidence**
    * use as many modes of evidence/displaying evidence as possible (modes of data presentation)
	* integrate words/numbers/images/diagrams (information rich)
	* analysis should drive the tool

* **Principle 5: Describe/document evidence with appropriate labels/scales/sources**
	* add credibility to that data graphic

* **Principle 6: Content is the most important**
	* analytical presentations ultimately stand/fall depending on quality/relevance/integrity of content

## Exploratory Graphs

* **Purpose**: What does my data look like? Understand data properties, find pattern in data, suggest modeling strategies (do we use linear or non-linear?), debug. 
* **Characteristics**: made quickly, large number produced, gain personal understanding, appearances and presentation are aren’t as important
    
### One Dimension Summary of Data
* `summary(data)` = returns min, 1st quartile, median, mean, 3rd quartile, max
* `boxplot(data$v1, col = “blue”)` = produces a box with middles 50% highlighted in the specified colour. Graphical output of `summary()`. 
* `histogram(data$v1, col = "green")` = produces a histogram with specified breaks and colour
    * `breaks = 100` = number of bars. The higher the number is the smaller/narrower the histogram columns are. Too big = noisey / rough, too small = can't see shape of distribution
* `rug(data$v1)` = density plot, adds strip under histrogram indicating location of each data point
* `barplot(table(data), col = "wheat", main = "Title")` = produces a bar graph, usually for categorical data

* **overlaying features**
    * `abline(h/v = 12)` = overlays a horizontal (boxplot) or vertical (histogram) line at specificed location. 
    * `col = “red”` = specifies color
	  * `lwd = 4` = line width
	  * `lty = 2` = line type

    * Horziontal line useful over `boxplot` to see how much data falls at specified value. 
      * ***example:*** How many islands speak 12 languages?
          * `boxplot(data)`;`abline(h = 12)`. If line is above box, then less than 75% (3rd quartile) of islands speak 12 languages.
    * Vertical line useful over `histogram` 
        * `abline(v = meadian(data), col = blue, lwd = 4)` = displays median line. ***note*** boxplot contains median line as feature, histogram does not
        
### Two Dimensional Summaries

* multiple/overlayed 1-D plots (Lattice/ggplot2 )
* **box plots**: `boxplot(pm25 ~ region, data = pollution, col = “red”)`   
* air pollution in east is higher than west. Lower in west but more spread out, with high extreme values

```{r fig.height = 3, fig.width = 4, fig.align='center', echo=FALSE}
grid.raster(readPNG("figures/3.jpg"))
```


* **histogram**:
    * `par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))` = set margin
    * `hist(subset(pollution, region == "east")$pm25, col = "green")` = first histogram
	  * `hist(subset(pollution, region == "west")$pm25, col = "green")` = second histogram
* air pollution in east is higher than west, but no extreme values. Lower in west but more spread out, with high extreme values


```{r fig.height = 3, fig.width = 4, fig.align='center', echo=FALSE}
grid.raster(readPNG("figures/4.jpg"))
```

* **scatterplot**
  * `with(pollution, plot(latitude, pm25))` = scatterplot of pollution going south to north (latitude). Plots two variables.
  * `abline(h = 12, lwd = 2, lty = 2)` = plots horizontal dotted line
  with(pollution, plot(latitude, pm25, col = region))` = same as first histogram, but data dots are coloured by region. Red = west, Black = East. Plots three variables.
  
```{r fig.height = 3, fig.width = 4, message = F, warning = F, fig.align='center', echo=FALSE}
grid.raster(readPNG("figures/5.jpg"))
```

* **mutiple scatterplot**
  * another way of looking at three variables instead of using colour
  * `par(mfrow = c(1, 2), mar = c(5, 4, 2, 1))` = sets margins
	* `with(subset(pollution, region == "west"), plot(latitude, pm25, main = "West"))` = left scatterplot
	* `with(subset(pollution, region == "east"), plot(latitude,  pm25, main = "East"))` = right scatterplot

```{r fig.height = 4, fig.width = 6, fig.align='center', echo=FALSE}
grid.raster(readPNG("figures/6.jpg"))
```
* both scatterplots (single & multiple) show that for pollution is higher in mid-latitudes than low / high latitudes for both eastern and western regions.

  * **scatterplot**
  * `with(pollution, plot(latitude, pm25, col = region))`
	* `abline(h = 12, lwd = 2, lty = 2)` = plots horizontal dotted line showing the standard air quality level
	* `plot(jitter(child, 4)~parent, galton)` = spreads out data points at the same position to simulate measurement error/make high frequency more visibble

* **summary**
  * quick and dirty  
  * Explore basic questions and hypotheses (perhaps rule them out)


### Process of Making a Plot/Considerations
* where will plot be made? screen or file?
* how will plot be used? viewing on screen/web browser/print/presentation?
* large amount of data vs few points?
* need to be able to dynamically resize?
* **plotting system**: base, lattice, ggplot2?

## Base Plotting
* blank canvas, “artist’s palette”, start with plot function
* annotations - text, lines, points, axis
* convenient, but cannot go back when started (need to plan ahead)
* everything need to be manually set carefully to be able to achieve the desired effect (margins)
* core plotting/graphics engine in R encapsulated in the following
    * ***graphics***: plotting functions for vase graphing system (plot, hist, boxplot, text)
	  * ***grDevices***: contains all the code implementing the various graphics devices (x11, PDF, PostScript, PNG, etc)

* ***Two phase***: initialize, annotate
    * calling `plot(x, y)` or `hist(x)` will launch a graphics device and draw a plot on device
    * if no argument specified, default called
	  * parameters documented in “`?par`"
	  * ***Note**: it is some times necessary to convert column/variable to factor to make plotting easier *
		  * `airquality <- transform(airquality, Month = factor(month))`

### Base Graphics Functions and Parameters
* **arguments**
    * `pch`: plotting symbol (default = open circle)
	  * `lty`: line type (default is solid)
		    * 0=blank, 1=solid (default), 2=dashed, 3=dotted, 4=dotdash, 5=longdash, 6=twodash
  	* `lwd`: line width (integer)
  	* `col`: plotting color (number string or hexcode, colors() returns vector of colors)
  	* `xlab`, `ylab`: x-y label character strings
  	* `cex`: numerical value giving the amount by which plotting text/symbols should be magnified relative to the default
		- `cex = 0.15 * variable`: plot size as an additional variable
* `par()` function = specifies global graphics parameters, affects all plots in an R session (can be overridden)
	  * `las`: orientation of axis labels
	  * `bg`: background color
  	* `mar`: margin size (order = bottom left top right)
  	* `oma`: outer margin size (default = 0 for all sides)
  	* `mfrow`: number of plots per row, column (plots are filled row-wise)
  	* `mfcol`: number of plots per row, column (plots are filled column-wise)
  	* can verify all above parameters by calling `par("parameter")`
* **plotting functions**
    * `plot` = make a scatterplot, or other plot depending on class of data
    * `lines` = add lines to existing plot. I.E. Connecting dots in a time-series plot
    * `points` = add points to a plot. I.E. add a different group or subset afterwards
    * `text` = add text labels within the plot using specified x,y coordinates 
    * `title` = add text labels outside the plot (x/y-axis, title, subtitle, outer margin)
    * `mtext` = add text to inner/outer margin of plot
    * `axis` = add axis ticks/labels

### Base Plot Example

```{r fig.height = 4, fig.width = 6, fig.align='center'}
library(datasets)
# type =“n” sets up the plot and does not fill it with data
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City"))
# subsets of data are plotted here using different colors
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
with(subset(airquality, Month != 5), points(Wind, Ozone, col = "red"))
legend("topright", pch = 1, col = c("blue", "red"), legend = c("May", "Other Months"))
# regression line is produced here
model <- lm(Ozone ~ Wind, airquality)
abline(model, lwd = 2)
```


### Multiple Plot Example
* ***Note**: typing `example(points)` in R will launch a demo of base plotting system and may provide some helpful tips on graphing *

```{r fig.height = 4, fig.width = 6, fig.align='center'}
library(datasets)
# `mfrow` sets three plots on a single device
# margins set a lil bit smaller than default
# `oma` sets outer margin to bigger than default so it can be filled with main title
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0))
with(airquality, {
  # here three plots are filled in with their respective titles
	plot(Wind, Ozone, main = "Ozone and Wind")
	plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
	plot(Temp, Ozone, main = "Ozone and Temperature")
	# this adds a line of text in the outer margin*
	mtext("Ozone and Weather in New York City", outer = TRUE)}
)
```

## Graphics Device
* A graphics device is something where you can make a plot appear
  * **window on screen** (screen device) = quick visualizations and exploratory analysis
	* **pdf** (file device) = plots that may be printed out or incorporated in to document
	* **PNG/JPEG** (file device) = plots that may be printed out or incorporated in to document
	* **scalable vector graphics** (SVG)
* When a plot is created in R, it has to be sent to a graphics device
* **Most common is screen device**
	* `quartz()` on Mac, `windows()` on Windows, `x11()` on Unix/Linux
	* `?Devices` = lists devices found
* **Plot creation**
	* screen device
		* call plot/xplot/qplot $\rightarrow$ plot appears on screen device $\rightarrow$ annotate as necessary $\rightarrow$ use
	* file devices
		* explicitly call graphics device $\rightarrow$ plotting function to make plot (write to file) $\rightarrow$ annotate as necessary $\rightarrow$ explicitly close graphics device with `dev.off()`
* **Graphics File Devices**
	* ***Vector Formats*** (good for line drawings/plots w/ solid colors, a modest number of points)
		* `pdf`: useful for line type graphics, resizes well, usually portable, not efficient if too many points
		* `svg`: XML based scalable vector graphics, support animation and interactivity, web based
		* `win.metafile`: Windows metafile format
		* `postscript`: older format, resizes well, usually portable, can create encapsulated postscript file, Windows often don’t have postscript viewer (postscript = predecessor of PDF)
	* ***Bitmap Formats*** (good for plots w/ large number of points, natural scenes/webbased plots)
		* `png`: Portable Network Graphics, good for line drawings/image with solid colors, uses lossless compression, most web browsers read this natively, good for plotting a lot of data points, does not resize well
		* `JPEG`: good for photographs/natural scenes/gradient colors, size efficient, uses lossy compression, good for plotting many points, does not resize well, can be read by almost any computer/browser, not great for line drawings (aliasing on edges)
		* `tiff`: common bitmap format supports lossless compression
		* `bmp`: native Windows bitmapped format
* **Multiple Open Graphics Devices**
	* possible to open multiple graphics devices (screen, file, or both)
	* plotting occurs only one device at a time
	* `dev.cur()` = returns the currently active device
	* every open graphics device is assigned an integer >= 2
	* `dev.set(<integer>)` = change the active graphics device
		`<integer>` = number associated with the graphics device you want to switch to
* **Copying plots**
	* `dev.copy()` = copy a plot from one device to another
	* `dev.copy2pdf()` = specifically for copying to PDF files
	* ***Note**: copying a plot is not an exact operation, so the result may not be identical to the original *
	* ***example***

```{r fig.height = 3, fig.width = 4, fig.align='center', results = 'hide'}
## Create plot on screen device
with(faithful, plot(eruptions, waiting))
## Add a main title
title(main = "Old Faithful Geyser data")
## Copy my plot to a PNG file
dev.copy(png, file = "geyserplot.png")
## Don't forget to close the PNG device!
dev.off()
```

## `lattice` Plotting System
* `library(lattice)` = load lattice system
* implemented using the `lattice` and `grid` packages
  * `lattice` package = contains code for producing ***Trellis*** graphics (independent from base graphics system)
	* `grid` package = implements the graphing system; lattice build on top of grid
* all plotting and annotation is done with ***single function call*** rather than piece by piece like `base plot`
	* margins/spacing/labels set automatically for entire plot, good for putting multiple on the screen
	* good for conditioning plots $\rightarrow$ examining same plots over different conditions how y changes vs x across different levels of z
	* `panel` functions can be specified/customized to modify the subplots
* lattice graphics functions return an object of class "trellis", where as base graphics functions plot data directly to graphics device
	* print methods for lattice functions actually plots the data on graphics device
	* trellis objects are auto-printed
	* `trellis.par.set()` $\rightarrow$ can be used to set global graphic parameters for all trellis objects
* hard to annotate, awkward to specify entire plot in one function call
* cannot add to plot once created, panel/subscript functions hard to prepare

### `lattice` Functions and Parameters
* **Funtions**
	* `xyplot()` = main function for creating scatterplots
	* `bwplot()` = box and whiskers plots (box plots)
	* `histogram()` = histograms
	* `stripplot()` = box plot with actual points
	* `dotplot()` = plot dots on "violin strings"
	* `splom()` = scatterplot matrix (like pairs() in base plotting system)
	* `levelplot()`/`contourplot()` = plotting image data
* **Arguments** for `xyplot(y ~ x | f * g, data, layout, panel)`
	* default blue open circles for data points
	* formula notation is used here (`~`) = left hand side is the y-axis variable, and the right hand side is the x-axis variable
	* `f`/`g` = conditioning/categorical variables (optional)
		* basically creates multi-panelled plots (for different factor levels)
		* `*` indicates interaction between two variables
		* intuitively, the xyplot displays a graph between x and y for every level of `f` and `g`
	* `data` = the data frame/list from which the variables should be looked up
		* if nothing is passed, the parent frame is used (searching for variables in the workspace)
		* if no other arguments are passed, defaults will be used
	* `layout` = specifies how the different plots will appear
		* `layout = c(5, 1)` = produces 5 subplots in a horizontal fashion
		* padding/spacing/margin automatically set
	* [optional] `panel` function can be added to control what is plotted inside each panel of the plot
		* `panel` functions receive x/y coordinates of the data points in their panel (along with any additional arguments)
		* `?panel.xyplot` = brings up documentation for the panel functions
		* ***Note**: no base plot functions can be used for lattice plots *

## ggplot2 system
* combines lattice and base 
* easier to use than lattice 
* good for XY-causation-plots.
